---
title: "DATA 606 Data Project Proposal"
author: "Thomas Hill"
output: pdf_document
---

### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# load data from the fivethirtyeight github
etoh_consumption <- read.csv('https://raw.githubusercontent.com/fivethirtyeight/data/master/alcohol-consumption/drinks.csv')

head(etoh_consumption)
names(etoh_consumption)

```

### Research question 

**You should phrase your research question in a way that matches up with the scope of inference your dataset allows for.**

What effect does alcohol type have on total alcohol consumption?


### Cases 

**What are the cases, and how many are there?**

The cases are average serving sizes per person for each country in 2010, distinguished by one of three alcohol types: beer, wine, and spirit.

### Data collection 

**Describe the method of data collection.**

These data were collected by the World Health Organization as part of its Global Information System on Alcohol and Health (GISAH). These are based on government records and metrics from alcohol companies to estimate consumption per country in 2010.  

### Type of study 

**What type of study is this (observational/experiment)?**

This study is observational as it aims to estimate the real consumption of alcohol in 2010. There is no randomization or treatment arm, the metrics are derived but taken as representative of the actual alcohol consumption. 

### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**

The raw data is available on GitHub <https://raw.githubusercontent.com/fivethirtyeight/data/master/alcohol-consumption/drinks.csv>, and the accompanying article was originally made available on <https://fivethirtyeight.com/features/dear-mona-followup-where-do-people-drink-the-most-beer-wine-and-spirits/>. More information is made available about the GISAH and its estimation methods on <https://www.who.int/substance_abuse/activities/gisah/en/>.

### Dependent Variable

**What is the response variable? Is it quantitative or qualitative?**

The response variables are the total alcohol consumption in liters per capita. It is a quantitative variable.

### Independent Variable

**You should have two independent variables, one quantitative and one qualitative.**

I'm looking primarily at the principle alcohol type (beer, wine, or spirit) to predict total alcohol consumption. I will also use the percent of total of each alcohol type as a quantitative measure of each country's trends. 

### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r total-alcohol-consumption}

summary(etoh_consumption$total_litres_of_pure_alcohol)
hist(etoh_consumption$total_litres_of_pure_alcohol)

```
  The total alcohol consumption per country is skewed to the right, likely because there are many countries where effective alcohol consumption is zero for religious reasons.  This is likely an underestimator of alcohol consumption in this country as alcohol per the GISAH methodology is based off governement and commercial resources, and alcohol has a long history of personal production.
  
```{r alcohol-type}

hist(etoh_consumption$beer_servings, xlim = c(0, 400), ylim = c(0,140))
hist(etoh_consumption$wine_servings, xlim = c(0, 400), ylim = c(0,140))
hist(etoh_consumption$spirit_servings, xlim = c(0, 400), ylim = c(0,140))
```

By serving, wine appears to be the most exclusive alcohol type. Beer and spirits seem to have small differences in consumption. For a final analysis, I'd like to see if there are any countries with majority- or plurality consumption trends, and in this case, if this could be used to predict total alcohol consumption.

Finally, for the last section I'll expresss the alcohol servings as a percent of total alcohol consumption. Since there are only the three sources of alcohol considered, I'll avoid any conversions from servings --> liters of pure alcohol by expressing the servings as a percent. 